---
title: "Nodes"
author: "Thomas Lin Pedersen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
set.seed(2022)
```

Nodes in a network are the entities that are connected. Sometimes these are also
referred to as vertices, but `ggraph` has opted for this nomenclature and uses it
consistently. While the nodes in a graph are the abstract concepts of entities, 
and the layout is their physical placement, the node geoms are the visual 
manifestation of the entities. Conceptually one can simply think of it in terms
of a scatter plot --- the layout provides the x and y coordinates, and these can
be used to draw nodes in different ways in the plotting window. Actually, due
to the design of `ggraph` the standard *scatterplot-like* geoms from `ggplot2`
can be used directly for plotting nodes:

```{r, message=FALSE}
library(ggraph)
library(tidygraph)

set_graph_style(plot_margin = margin(1,1,1,1))
gr <- as_tbl_graph(highschool)

ggraph(gr, layout = 'kk') + 
  geom_point(aes(x = x, y = y))
```

The reason this works is that, as discussed in the Layout vignette, layouts 
return a `data.frame` of node positions and metadata and this is used as the 
default plot data:

```{r}
head(create_layout(gr, layout = 'kk'))
```

## `geom_node_*()`
While usage of the default `ggplot2` is absolutely allowed, `ggraph` comes with
its own set of node geoms. Many of these are direct translations of `ggplot2`
own geoms like `geom_point()` so one could wonder why bother to use them. 

The first reason is to provide clear code. It is not apparent anywhere that the
standard geoms are addressing the nodes and using `geom_node_*()` makes it clear
that this layer will draw nodes. 

The second reason is that it will save typing.
Since `ggraph` is in control of the shape of the input data through the layout
calculations, it knows that *x* and *y* position is encoded in an `x` and `y` 
column. This means that `geom_node_*` can default the x and y aesthetics so 
there's no need to type them:

```{r}
ggraph(gr, layout = 'kk') + 
  geom_node_point()
```

sometimes there is a need for addressing the x and y aesthetics, which is still
possible, for instance if a partition layout should be inverted:

```{r}
gr <- tbl_graph(flare$vertices, flare$edges)

ggraph(gr, layout = 'partition') + 
  geom_node_tile(aes(y = -y, fill = depth))
```

of course this could also be accomplished by reversing the y-axis using 
`scale_y_reverse()` so this is just to illustrate that the defaults are easily
overwritten if needed.

The third reason is for the added functionality. All `ggraph` geoms get a 
`filter` aesthetic that allows you to quickly filter the input data. The use of
this can be illustrated when plotting a tree:

```{r}
ggraph(gr, layout = 'dendrogram', circular = TRUE) + 
  geom_edge_diagonal() + 
  geom_node_point(aes(filter = leaf)) + 
  coord_fixed()
```

In the above plot only the terminal nodes are drawn by filtering on the logical
leaf column provided by the dendrogram layout.

## The different node geoms
The usual suspects are of course provided in the form of `geom_node_point()` 
(showcased above), `geom_node_text()`, and `geom_node_label()`. These work as
expected, taking in the usual aesthetics (plus *filter*). Only x and y are
defaulted so everything else must be provided e.g. label which does not default
to the `name` column like is done in `igraph`. One feature sets 
`geom_node_text()` and `geom_node_label()` apart from their `ggplot2` 
counterparts: both have a `repel` argument that, when set to `TRUE`, will use 
the repel functionality provided by the 
[ggrepel](https://CRAN.R-project.org/package=ggrepel) package to avoid 
overlapping text. There is also `geom_node_voronoi()` that plots nodes as cells
from a voronoi tesselation. This is useful for e.g. showing dominance of certain
node types in an area as overlapping is avoided:

```{r}
graph <- create_notable('meredith') %>% 
  mutate(group = sample(c('A', 'B'), n(), TRUE))

ggraph(graph, 'stress') + 
  geom_node_voronoi(aes(fill = group), max.radius = 1) + 
  geom_node_point() + 
  geom_edge_link() + 
  coord_fixed()
```

Apart from these geoms there's a set of geoms mainly useful for spatial
node layouts such as treemaps, partition, circle packing, and fabric. 
`geom_node_tile()` and `geom_node_range()` are the `ggraph` counterpart to 
`ggplot2`s `geom_tile()` and `geom_linerange()` while 
`geom_node_circle()` and `geom_node_arc_bar()` maps to `ggforce`s 
`geom_circle()` and `geom_arc_bar()`. Collective for these is that the spatial
dimensions of the geoms (e.g. radius, width, and height) are precalculated by
their intended layouts and defaulted by the geoms:

```{r}
ggraph(gr, layout = 'treemap', weight = size) + 
  geom_node_tile(aes(fill = depth))
```

All spatial node geoms will be center-based, meaning that the x and y value of
the layout will refer to the center of the layout and not e.g. the bottom-left
corner. This makes it easier to add labels to spatial layouts as well as using
spatial layouts in a non-spatial way:

```{r}
l <- ggraph(gr, layout = 'partition', circular = TRUE)
l + geom_node_arc_bar(aes(fill = depth)) + 
  coord_fixed()
```

```{r}
l + geom_edge_diagonal() + 
  geom_node_point(aes(colour = depth)) + 
  coord_fixed()
```

More node geoms are sure to appear in `ggraph` with time but they will generally
be quite easily comprehensible due to their strong resemblance to the standard
`ggplot2` geoms. After all it is just points on a plane...

## Want more?
Check out the other vignettes for more information on how to specify 
[layouts](Layouts.html) and draw [edges](Edges.html)...
